The Markov Link Method: a nonparametric approach to combine observations from multiple experiments

### 2018-08-06 paper/paper.tex
Confidence intervals, which we're now ignoring...

\newcommand{\figgygr}[2]{\includegraphics[width=#1\textwidth]{#2}}
\newcommand{\figgygrs}[3]{\begin{tabular}{c}\includegraphics[width=#1\textwidth]{#2}\\#3\end{tabular}}
\newcommand{\figgygrsTable}[3]{\begin{tabular}{c}{#2}\\#3\end{tabular}}
\newcommand{\figgygrsH}[3]{\begin{tabular}{c}\includegraphics[width=#1\textwidth,angle=270]{#2}\\#3\end{tabular}}

% \newcommand{\figgygr}[2]{pic}
% \newcommand{\figgygrs}[3]{\begin{tabular}{c}pic\\#3\end{tabular}}
% \newcommand{\figgygrsTable}[3]{\begin{tabular}{c}{#2}\\#3\end{tabular}}
% \newcommand{\figgygrsH}[3]{\begin{tabular}{c}pic\\#3\end{tabular}}

@@

\title{Markov Link Method for calibrating without joint measurement, including the case of destructive measurements}
\author{Jackson Loper, Osnat Penn, Trygve Bakken, David Blei, Liam Paninski}

@@

\begin{abstract}
A proliferation of new experimental tools has left a serious gap: calibration. Two thermometers can be calibrated against each other by simply measuring the same bodies of water with both thermometers, but the problem is much harder for many modern tools.  One common problem is that we do not have measurements from the same ``body of water'' for both tools.  We propose the Markov Link Method (MLM) as a way to overcome this difficulty.   This method produces consistent estimators that tightly bound the calibration, i.e. the conditional distribution of one tool's measurement given another tool's measurement. It achieves this without any measurement data from both tools applied to the same ``bodies of water.'' Moreover, MLM can be applied even if we cannot make any assumptions about what calibrations we might expect to see; instead, it uses a subpopulation-based conditional independence assumption.   We evaluate MLM on a pair of single-cell RNA techniques, obtaining a calibration between the tools.
\end{abstract}

The modern setting is rife with experimental measurement tools, and it can be very frustrating to understand how the output of these tools relate to one another.  This problem is known as ``calibration'' or ``zeroing.''  A calibration tells us what readings we should expect from one tool, given the reading we obtained from another tool.   Calibration additionally must give uncertainty bounds for how much we can trust those expectations  \cite{bipm2008international}.  Calibration between measurement tools allows us to combine experimental results from different labs and different methodologies into larger scientific theories.

Formally, a calibration is simply a conditional distribution.  We will denote it by $q^*(y|x)$.  Let us say we obtain measurement result $x$ from one tool on a particular specimen.  The number $q^*(y|x)$ indicates the probability of obtaining result $y$ from a second tool applied to measure the same specimen.  One way to learn the calibration is to measure the same specimens with both tools.  We call this ``joint measurement.''  Unfortunately, calibrations are often required even when joint measurement is unavailable.  For example, if the measurement tool significantly alters the specimen being measured, joint measurement is simply impossible.  In other cases, it may be expensive or impractical.

We here propose the Markov Link Method (MLM) to estimate calibrations between tools.  The MLM can be applied without any joint measurement.  The key idea is to use multiple subpopulations of specimens.  If each subpopulation captures a different slice of the overall population, we can obtain tight bounds on the true calibration.  This is true even if the subpopulations are highly heterogeneous and overlapping.  By integrating information from all the subpopulations we can make rigorous deductions about what the calibration might be.  MLM also gives suggestions about which further subpopulations might be helpful to study in order to further refine our knowledge of the true calibration.

\figgy{
\hfill{}\figgygrs{.5}{../images/bigpicture}{(a) Setup: each specimen can be \\ measured with either tool I or tool II, \\ but can't be measured with both.\\ How do we calibrate the tools?}
\hfill{}\figgygrs{.4}{../images/plate}{(b) Key modeling assumption: \\ the calibration $q$ is the same \\ for every subpopulation. }\hfill{}

\caption{{\bf The Markov Link Method assumption makes calibration possible, even without joint measurement}.  In subfigure (a) we measure some neurons with tool I and others with tool II.  How can we calibrate the tools?  That is, how can we know what we might have seen if we had measured the first group with tool II or the second group with tool I?  The Markov Link Method (MLM) gives a way to get a partial answer, using a subpopulation analysis.  In subfigure (b) we present a plate diagram describing our beliefs about the conditional independencies in this situation; cf.\ \cite{koller2009probabilistic} for a thorough explanation of this type of diagram.  For each specimen we are interested in $\ell$ (subpopulation), $X$ (tool-I measurement), and $Y$ (tool-II measurement).  For $n$ samples we can see $(\ell,X)$ and for $m$ samples we can see $(\ell,Y)$.  However, for the first group of $n$ samples we cannot observe $Y$, and for the second group of $m$ samples we canot observe $X$.  This is indicated in the diagram by the grayed-out circles.   This lack of joint measurement makes it hard to learn the variable distributions.  The MLM has a key assumption that makes it possible: we assume $X$ is sufficiently detailed to \emph{statistically isolate} $Y$ from $\ell$.  This assumption is reflected in the arrows of the diagram; all paths from $\ell$ variables to $Y$ variables pass through $X$ variables.  The assumption allows us to learn much about the important probability distributions, as parametrized by the tool-I measurement distribution $p(x|\ell)$ and the calibration $q(y|x)$. \label{fig:bigpicture}}
}

This paper will proceed in the following sections:
\begin{itemize}
    \item Description of how to apply the Markov Link Method, using an example to explain the process
    \item A variety of examples where the method may apply
    \item The method's performance in simulation
    \item The method's results on real-world data
    \item Relation to prior work
    \item Discussion
\end{itemize}

@@

We begin with an example.

Imagine that we have a large collection of human cells.  We will call each cell a `specimen.'  For each specimen we have:
%
\begin{enumerate}
    \item $\ell$, the sampling strategy used to obtain the specimen.  We will also call $\ell$ the specimen ``subpopulation,'' for each strategy should capture a different subpopulation of the overall set of specimens.  For example, we could have a few strategies based on selecting cells of particular sizes, a few strategies based on selecting cells from particular places in the body, and a few strategies based on based on selecting cells with particular proteomic markers.  Each additional sampling strategy shows us a different subpopulation of the cells and strengthens what kinds of inferences are possible. 
    \item $X$, the cell's `transcriptomic type,' as measured by looking at the transcriptomic activity in the cell.  
    \item $Y$, the cell's `morphological type,' as measured by looking at an image of the cell.
\end{enumerate}
%
Our task is to reconcile the two different notions of `cell-type' indicated by $X$ and $Y$, by constructing what we call a `calibration.'  This calibration should tell us what transcriptomic types are associated with what morphological types.  For example, we might like to be able to say ``This cell has transcriptomic type 3, therefore it probably has morphological type F.''  We can make math of this idea by defining a calibration as a conditional probability $q^*(y|x,\ell)$. This object indicate the likelihood that the morphological type is $y$ given that the transcriptomic type is $x$ and the specimen was sampled with strategy $\ell$.  In these terms, our task is to learn the calibration $q^*$.

Learning $q^*$ is hard when ``joint measurement'' is impossible or impractical.  By this we mean that for any given specimen we can either observe $\ell, X$ or $\ell,Y$.  We can never observe $\ell,X,Y$ for any specimen.  This makes it much more difficult to estimate the calibration; the fundamental difficulty is sketched in Figure \ref{fig:bigpicture}.  To make calibration possible without joint measurement, we need an assumption additional assumption.  The Markov Link Method supplies this assumption in the form of a conditional independence assumption.  This makes it possible to learn about the calibration.

To make it possible to learn $q^*$ even without joint measurement, the MLM makes the assumption that the calibration is the same for every sampling strategy:

\begin{center}
\fbox{\parbox[c]{4in}{
\vspace{.1in}
\begin{center}
\textbf{The Markov Link Method Assumption}\\$q^*(y|x,\ell)=q^*(y|x)$ for each $\ell$
\end{center}
\vspace{.1in}
}}
\end{center}

This assumption can be understood through the following thought experiment.  Let us say we are told that a cell has transcriptomic type $x$.  We could then make predictions about the morphological type $y$.  Now we are told new information: the cell was obtained via sampling strategy $\ell$.  Does this change our predictions about the morphological type?  If so, the MLM assumption is violated.  However, if $X$ is sufficiently informative then learning the sampling strategy will not change our predictions about the morphological type.  In this case, the MLM assumption holds.

If the MLM assumption holds, a frequentist point of view suggests the following procedure for estimating the calibration $q^*$ from data:

\figgy{
\hfill{}\figgygr{.9}{../images/qtildesamps}\hfill{}
\caption{{\bf Estimating the calibration and measuring our uncertainty}.  \label{fig:qtildesamps} On the left we visualize our best guess for true the calibration between two different tools for measuring cell-type.  Tool I uses one approach to obtain a fine-grained classification of cells, whereas tool II uses a different approach and only gives coarse type indication.  We can visualize our estimated calibration with a heatmap, formed by drawing a rectangle for each $x,y$ with a color indicating the magnitude of $\tilde q(y|x)$.  For example, examine the bright yellow square on the `Pvalb Gpr149' row and the `unknown' column.  This suggests that if a cell has the `Pvalb Gpr149' type according to tool I, it is extremely likely to be assigned the `unknown' type by tool II.  However, our best guess might not be correct.  On the right we show many plausible alternatives for the `Pvalb Gpr149' row of the calibration.  Each possibility is drawn from a 95\% confidence region for calibration.  Some of the alternatives suggest that some `Pvalb Gpr149' cells might not be classified as `unknown,' but might instead be correctly identified as being of the `Pvalb' type.  The data we have cannot tell us which of these alternatives is closest to the truth.  More data and more subpopulations would be necessary to reduce the variability within this confidence region.  However, we can have some confidence that `Pvalb Gpr149' cells are not being classified as `Meis2' or `Vip', insofar as none of the plausible calibrations give high probability to these tool II types.}
}

\begin{enumerate}
    \item Summarize our observed data in two matrices:
    \begin{align*}
    N^X_{\ell x} &= |\{i\leq n:\ \ell_i=\ell,X_i=x\}|\\
    N^Y_{\ell y} &= |\{j\leq m:\ \ell_j=\ell,Y_j=x\}|
    \end{align*}
    That is, $N^X_{\ell x}$ indicates the number of specimens from subpopulation $\ell$ with transcriptomic type $x$.  Likewise $N^Y_{\ell y}$ indicates the number of specimens from subpopulation $\ell$ which with morphological type $y$.

    \item Define a confidence region for calibration $q^*$.  To do this we incorporate a nuisance parameter: let $p^*(x|\ell)$ denote the probability that a cell sampled with strategy $\ell$ will have transcriptomic type $x$.  We form a joint simultaneous confidence region for all of $p^*,q^*$ using the following likelihood ratio statistic:
    \[
    L(p,q)=\frac{1}{\Omega_X}\sum_{\ell,x} N^X_{\ell x} \log \frac{\hat p(x|\ell)}{p(x|\ell)} + \frac{1}{\Omega_Y}\sum_{\ell,y} N^Y_{\ell y} \log \sum_x \frac{\hat h(y|\ell)}{p(x|\ell) q(y|x)}
    \]
    where $\hat p(y|x)=N^X_{\ell x} / \sum_{x'}N^X_{\ell x'}$ and $\hat h(y|x)=N^Y_{\ell y} / \sum_{y'}N^Y_{\ell y'}$ denote empirical conditional distributions.  We then define our confidence region as 
    \[
    R = \{(p,q):\ L(p,q) < k\}
    \]
    and choose $k$ to ensure any desired coverage.  For example, to produce a 95\% confidence region, we would like to choose $k$ so that the probability that $p^*,q^* \in R$ is at least 95\%.  We refer the reader to Appendix \ref{sec:confinterval} for details on choosing $k$.

    \item Determine whether the confidence region is empty.  If the region is empty then we can reject the MLM assumption hypothesis with a significance level of one minus the region coverage.  Assuming we fail to reject, we can continue:

    \item Sketch out the the confidence region, using examples.  In most nontrivial cases the region $R$ is high-dimensional and difficult to inspect directly.  We therefore produce examples from this region to try to help the user understand what values of $q$ are plausible.  In particular, we sample $Z_{x,y} \sim \mathrm{Exponential}(1)$ for each $x,y$ and solve the problem
    \[
    \max_{p,q \in R} \quad \sum_{x,y} Z_{x,y} \log q(y|x) \\
    \]
    This problem does not have a closed-form solution, but can generally be solved using an iterative procedure; we describe our method in detail in Appendix \ref{sec:optproblems}.  The result is a particular value of $q$ which lies within $R$.  In practice, we find these examples are almost always on the boundary of $R$.  We can then repeat this process by resampling $Z$.  By repeating this process many times, we can get a large collection of plausible values of $q$.  For any fixed value of $x$ we can visualize these samples by using a heatmap to plot $q(\cdot |x)$ for all the samples.  See Figure \ref{fig:qtildesamps} for an example.

    \item Produce a point estimate.  It is not generally possible to produce a consistent point estimate.  Indeed, there are many situations in which the confidence region $R$ does not concentrate even in the limit of infinite data.  Nonetheless, if a point estimate is desired, one may pick point estimates $\tilde p,\tilde q$ by solving the minimization problem $\min L(p,q)$.  We caution that this problem does not always have a unique solution, so the result may depend upon the minimization strategy used.  Precise details for our algorithm may be found in Appendix \ref{sec:optproblems}.
\end{enumerate}

A Bayesian point of view would suggest a different procedure; this point of view may provide a useful complementary perspective and is worthy of future research.  However, we caution that it is nontrivial to choose reasonable prior distributions for $p^*,q^*$.  Priors which may seem uninformative (such as the Dirichlet distribution) can actually carry significant and unanticipated weight, due to the high-dimensional nature of the problem.  We give an example in \ref{sec:bayesian}.

The frequentist confidence region allows us to avoid making any modeling assumptions.  However, we are then faced with the task of understanding a high-dimensional confidence region.  As described above, our solution is to look at examples from inside the region.  We emphasize that these examples are not drawn uniformly within the confidence region; instead the examples illustrate something more like the boundary of the region, sketching out what might be possible.  We concede that this method is somewhat ad-hoc and hope that future research may discover a superior method to visualize this high-dimensional confidence region.  For now, it appears to be a practical solution.

@@

    \item Combining knowledge across experimental modalities: morphology and transcriptomics.  There are different ways to think about the different types of cells in an organism.  A traditional approach is to classify cells based on what they look like (cf.\ \cite{steinman1973identification,bloomfield1982physiological}).  A more modern approach is to assay the cell's transcriptome (cf.\ \citep{tasic2017shared}).  Unfortunately, modern high-resolution cell photography and single-cell sequencing technologies are both destructive.  As a result, we can't always get both kinds of data for the same specimens.  For cells native to regions full of diverse cell-types, it is thus quite hard to grasp the correspondence between these different kinds of classification systems.  The result is two completely independent classifications of cells, one for each way of looking at the cell.  MLM allows us to estimate the relationship between those two classification systems, yielding a holistic understanding of the different types of cells.  In this case, $\ell$ might indicate some side information such as where in the body the cell was found, $X$ would indicate a detailed classification of the cell according to its transcriptomics, and $Y$ would indicate a coarser classification of a cell according to its morphology.  We expect that cell morphology is largely a function of cell transcriptomics.  Thus, as long as the $X$ measurement is sufficiently detailed, we expect that any correlations between $Y$ and $\ell$ would be explained by $X$.  That is, the MLM assumption holds.

@@

\section{Simulation results}

@@

We use simulations to test the capabilities of the Markov Link Method to estimate a calibration.  We will:

\begin{itemize}
    \item Pick parameters to be `ground truth': 
    \begin{itemize}
        \item A calibration, $q^*(y|x)$.  
        \item A tool-I measurement distribution, $p^*(x|\ell)$.  Recall that this indicates the probability that tool I measures $x$ for a specimen gathered with sampling strategy $\ell$. 
        \item A tool-II measurement distribution, $h^*(y|\ell)$.   This indicates the probability that tool II measures $y$ for a specimen gathered with sampling strategy $\ell$.  We assume the the MLM assumption holds, and so this can be calculated from $q^*,p^*$ by the formula $h^*(y|\ell)=\sum_x p^*(x|\ell)q(y|x)$.  
    \end{itemize}
    \item Select a number of samples to take per subpopulation, $s$, and produce a simulated dataset for each subpopulation $\ell$:
    \begin{align*}
    (N^X_{\ell 1},N^X_{\ell 2}\cdots N^X_{\ell \Omega_X}) \sim \mathrm{Multinomial}(s,p^*(\cdot | \ell))\\
    (N^Y_{\ell 1},N^Y_{\ell 2}\cdots N^Y_{\ell \Omega_Y}) \sim \mathrm{Multinomial}(s,h^*(\cdot | \ell))
    \end{align*}
    \item Apply the Markov Link Method to the generated samples to obtain a calibration estimate $\tilde q$.
    \item Compute the error of the estimate $\tilde q$, as measured by a kind of averaged total variation distance, namely
    \[
    \frac{1}{2\Omega_{\ell}} \sum_{\ell x} |\tilde q(y|x) - q^*(y|x)|
    \]
    This distance ranges between zero and one.  A distance of zero indicates that $\tilde q=q^*$, and a distance of one indicates that the two calibrations place mass on completely disjoint sets, i.e. $q(y|x)=0$ whenever $\tilde q(y|x)>0$ and vice-versa.     
\end{itemize}

\figgy{
\hfill{}\figgygrs{.4}{../images/simulationIandIIpq.png}{(a) `Ground truth' for $p^*,q^*$.}\hfill{}
\hfill{}\figgygrs{.5}{../images/simulationsIandII.png}{(b) MLM performance \\ in estimating $q^*$.}\hfill{}

\caption{{\bf The Markov Link Method accurately estimates the calibration, as long as there are enough subpopulations}.  \label{fig:simsIandII}  We pick parameters as `ground truth' for $p^*,q^*$; these are as shown in subfigure (a) as heatmaps.  We then create simulated datasets using these parameter values, use the Markov Link Method to produce its estimator, and see and measure the total variation error of the estimator.  This error ranges from zero to one, where zero indicates no error and one indicates complete disagreement.  In subfigure (b) we see that the MLM is able to estimate the ground truth with negligible error, as long as it has enough samples.  However, we also consider another simulation: we merge some of the subpopulations in the ground truth parameters so that they become indistingiushable. In this degraded setup, the MLM is unable to learn the true calibration no matter how many samples we have. }
}

For the first simulation, we will look at a case in which the MLM performs well.  We will assume there are twenty different samping strategies, and tools I and II both measure specimens as being in one of six categories.  Our particular choices of ground truth can be found in Figure \ref{fig:simsIandII}.    This figure also shows that the MLM is able to accurately estimate the ground truth calibration $q^*$, as long as there are enough samples.

>>>>

% \newcommand{\figgygr}[2]{\includegraphics[width=#1\textwidth]{#2}}
% \newcommand{\figgygrs}[3]{\begin{tabular}{c}\includegraphics[width=#1\textwidth]{#2}\\#3\end{tabular}}
% \newcommand{\figgygrsTable}[3]{\begin{tabular}{c}{#2}\\#3\end{tabular}}
% \newcommand{\figgygrsH}[3]{\begin{tabular}{c}\includegraphics[width=#1\textwidth,angle=270]{#2}\\#3\end{tabular}}

\newcommand{\figgygr}[2]{pic}
\newcommand{\figgygrs}[3]{\begin{tabular}{c}pic\\#3\end{tabular}}
\newcommand{\figgygrsTable}[3]{\begin{tabular}{c}{#2}\\#3\end{tabular}}
\newcommand{\figgygrsH}[3]{\begin{tabular}{c}pic\\#3\end{tabular}}

@@

\title{The Markov Link Method: a nonparametric approach to combine observations from multiple experiments}
\author{Jackson Loper, Osnat Penn, Trygve Bakken, David Blei, Liam Paninski}

@@

\begin{abstract}
Using observations from multiple experiments is essential to building larger scientific theories.  For example, we might analyze different aspects of human neurons, such as morphology, transcriptomics, or electrophysiology.  A strong theory would account for each perspective as well as the relationship between the perspectives.  Unfortunately, we rarely have a single experiment that simultaneously measures every viewpoint, uses a sufficient sample size, and gets sufficient measurement precision.  In practice, we must therefore conduct different experiments and find a way to link the results.  The Markov Link Method (MLM) offers an algorithm to reason about these `links.' The method is nonparametric: it makes no assumptions about what the link should look like.  We evaluate the MLM on a pair of single-cell RNA techniques and gain new insight about the link between the two methods.
\end{abstract}

The last century has seen a proliferation of experimental techniques.   Different perspectives arise from different techniques (and even from different labs performing the `same' protocol).  Ideally, this yields theories that account for all perspectives.  Instead, sometimes, this leaves us with many perspectives and no idea how they are connected.

Our goal is to connect different perspectives.  To make this idea rigorous, we assume each experiment has two aspects:
%
\begin{itemize}
    \item a \emph{sampling procedure} determines what we study
    \item a \emph{measurement procedure} defines the perspective we take in studying it.
\end{itemize}
%
If two experiments differ in both aspects, it may be difficult to gain insight by combining their data.  For example, imagine one experiment performs a transcriptomic measurement of liver cells and another applies a morphological measurement of heart cells.  Without additional experiments or knowledge, looking at these two experiments together may not be very helpful.  On the other hand, if two experiments have some aspect in common, there is something to be learned.  In this sense, a collection of experiments can be understood as a network: a pair of experiments are connected if they have some aspect in common.  In some cases, we can also connect two experiments if they differ in both aspects but we have prior knowledge (e.g.\ if the sampling procedures are different, but a reweighting can be used to correct for the difference).    When many experiments can be joined into a common network, much can be learned.

We here propose the The Markov Link Method (MLM), an algorithm that uses a network of experiments to learn the relationship between two different perspectives.  We formalize this through what we call the `measurement link' parameters, $q^*(y|x)$.  The meaning of these parameters is defined as follows.  Let us say we obtain measurement result $x$ from one measurement procedure.  The number $q^*(y|x)$ indicates the probability of obtaining result $y$ from the second measurement procedure applied to measure the same specimen.  The link tells us how the properties measured by each procedure are associated.  For example, the link lets us answer questions such as ``I have observed the transcriptome of this cell -- what morphologies might this cell have?''

If we have a single large experiment that measures each specimen with both measurement procedures, we can estimate the link without the MLM.  In this case it is straightforward to study the link between the two perspectives, because we can examine specimens through both perspectives simultaneously.  However, in many cases we have some experiments in which specimens are measured with one procedure, other experiments using the other procedure, and few or no experiments where specimens are measured with both procedures.  This is particularly true if the measurement techniques destroy or alter the specimen in the process of measuring it.

In this paper we focus particularly on the case that we have a network of experiments based on several sampling procedures and exactly two different measurement procedures.  We also suppose that the measurement procedures return categorical observations, i.e.\ a `measurement' assigns a specimen to one of a finite set of categories.  The MLM could also generalize to numerical observations, but in that case it would be appropriate to incorporate some knowledge of smoothness (e.g.\ through a Gaussian assumption).  Categorical data requires no additional modeling assumptions, so for simplicity we will here focus on the categorical case.  This setup, sketched in Figure \ref{fig:bigpicture}, is perhaps the simplest example that highlights all the important assumptions and properties of the MLM.

\figgy{
\hfill{}\figgygrs{.5}{../images/bigpicture}{(a) Setup: each specimen can be \\ measured with one of two procedures, \\ but can't be measured with both.\\ How do we study the relationship \\ between the procedures?}
\hfill{}\figgygrs{.4}{../images/plate}{(b) Key modeling assumption: \\ the link $q$ is the same for \\ every sampling procedure. }\hfill{}

\caption{{\bf The Markov Link Method (MLM) links different measurement procedures by using a set of different sampling procedures.}  In subfigure (a) we have neurons gathered using many different sampling procedures.  We select half of the neurons and determine what type of neuron they are, using transcriptomic measurements.  We classify the second half of neurons based on a morphological properties.  The MLM gives a way to link these two types of measurement and reconcile the two notions of cell-type.  In subfigure (b) we present a plate diagram articulating the statistical assumption that makes it possible (cf.\ \cite{koller2009probabilistic} for a thorough explanation of this type of diagram).  For each neuron in each group, this diagram considers the sampling procedure ($\ell$), the transcriptomic type ($X$), and the morphological type ($Y$).  The assumption is that the transcriptomic type is sufficiently detailed to \emph{statistically isolate} the transcriptomic type from the sampling procedure.  This is reflected in the arrows of the diagram; all paths from $\ell$ variables to $Y$ variables pass through $X$ variables.  The assumption allows us to estimate the link, $q$. \label{fig:bigpicture}}
}

@@

Imagine that we have a large collection of human cells.  For each cell we are interested in 
%
\begin{enumerate}
    \item $\ell$, the sampling procedure used to obtain the specimen.  For example, we can use select different subpopulations of cells based on proteomic markers.  For each marker we might have a different sampling procedure designed to select cells with that marker.
    \item $X$, the cell's `transcriptomic type,' as measured by looking at the transcriptomic activity in the cell.  
    \item $Y$, the cell's `morphological type,' as measured by looking at an image of the cell.
\end{enumerate}
%
Our task is to reconcile the two different notions of `cell-type' indicated by $X$ and $Y$.  We would like to know what transcriptomic types are associated with what morphological types.  For example, we might like to say ``This cell has transcriptomic type 3, therefore it probably has morphological type F.''  Our main assumption is that these kinds of associations are unaffected by the choice of sampling procedure.  In particular:

\begin{center}
\fbox{\parbox[c]{4in}{
\vspace{.1in}
\begin{center}
\textbf{The Markov Link Method Assumption}\\$\mathbb{P}(Y|X,\ell)=q^*(y|x)$ for each $\ell$
\end{center}
\vspace{.1in}
}}
\end{center}

Here $\mathbb{P}(Y|X,\ell)$ indicates the probability that the morphological type is $y$ given that the transcriptomic type is $x$ and the specimen was sampled with strategy $\ell$.  The assumption is that this probability actually does not depend upon $\ell$.  This can be understood intuitively through a thought experiment.  Let us imagine we have attained perfect understanding of cellular biology.  Someone selects a cell using one of a set of sampling procedures and measures the transcriptomic activity of the cell.  We are then told the cell's transcriptomic activity (but we are not told the sampling procedure used to gather the specimen).  Using our perfect knowledge of the physical systems, we could then make predictions about a measurement of the cell's morphology.  Now we are told new information: we are told the sampling procedure used to gather the specimen.  Would this substantially change our predictions?  If so, the MLM assumption is violated.  However, if the transcriptomic measurement is sufficiently informative, learning the sampling strategy would not substantially change our predictions about the cell morphology.  As long as we can find a measurement procedure that is sufficiently informative in this way, we can apply the MLM.  A list of examples where the MLM might apply may be found in Section \ref{sec:examples}.

Our task is to estimate the link, $q^*$, from a collection of experiments.  We suppose that for each sampling procedure we have two experiments: one that measured transcriptomics and another that measured morphology.  We have no experiments in which both types were measured for the same specimens.  This setup is sketched in Figure \ref{fig:bigpicture}.  Different setups could also be considered.   For example, we might additionally have a small experiment in which both measurement procedures could be applied to the same specimen.  We might have more than two measurement procedures.  We might have measurement procedures that yield numerical observations instead of categorical ones.  Generalizing the method to all these cases should be straightforward, but we leave it for future work.

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $N^X$ and $N^Y$.  Here $N^X_{\ell x}$ indicates the number of specimens gathered with procedure $\ell$ that have transcriptomic type $x$.  Likewise $N^Y_{\ell y}$ indicates the number of specimens gathered with procedure $\ell$ that have morphological type $y$.

Taking a frequentist point of view, the Markov Link Method uses this data to produce two kinds of objects:
\begin{itemize}
    \item $\tilde q(y|x)$, a point estimate for the link parameter $q^*(y|x)$.  To obtain this we must incorporate nuisance parameters: let $p^*(x|\ell)$ denote the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$.  The log likelihood of the data is then given by
    \[
    L(p,q)=\left(\sum_{\ell,x} N^X_{\ell x} \log p(x|\ell) + \sum_{\ell,y} N^Y_{\ell y} \log \left(\sum_x p(x|\ell) q(y|x)\right)\right)
    \]
    We use a maximum likelihood approach to build our point estimators:
    \[
    \tilde p,\tilde q \in \argmax_{p,q} L(p,q)
    \]
    We caution that this optimization problem does not always have a unique maximizer, so the result may depend upon the maximization strategy used.  Precise details for our optimization algorithms may be found in Appendix \ref{sec:optproblems}.  This non-uniqueness is related certain so-called `identifiability' issues: in some cases it may be impossible to obtain a perfect point estimate even with an infinite number of samples.   We discuss this issue at length in Appendix \ref{sec:casestudies}.  In light of these concerns, it is particularly important to also consder our uncertainty:
    %
    \item $C_{x,y}$, a 95\% confidence interval for $q^*(y|x)$.  We obtain these with profile likelihood approach.  The profile likelihood is defined as
    \[
    L_{x y}(\xi)=\max_{\substack{p,q \\ q(y|x)=\xi}} L(p,q)
    \]
    This is the maximum possible likelihood of the data under the constraint that $q(y| x)=\xi$.  We then take $L^* = \sup_{p,q} L(p,q)$ and define a confidence region for $q^*( y| x)$ by
    \[
    R_{x y} = \left\{\xi:\ L^* - L(\xi) < k\right\}
    \]
    We choose $k$ to achieve the desired coverage.  That is, we would like to choose $k$ so that $q^*(y|x) \in R_{x y}$ with probability at least 95\%.  We refer the reader to Appendix \ref{sec:confinterval} for details on choosing $k$.  Other coverages besides 95\% can also be produced by the same method.  Finally, for simplicity of visualization we transform the set $R_{x y}$ (which could be composed of several disjoint intervals) into a simple interval by taking $C_{x,y} = [\min R_{x,y},\max R_{x,y}]$.  Since this interval is a superset of the region, it will certainly contain $q^*(y|x)$ as long the region does.
\end{itemize}

A Bayesian point of view would suggest a different procedure, one which might provide a useful complementary perspective.  This is worthy of future research.  However, we caution that it is nontrivial to choose reasonable prior distributions for $p^*,q^*$.  Priors which may seem uninformative (such as the Dirichlet distribution) can have significant unanticipated consequences, due to the high-dimensional nature of the problem.  We give an example in \ref{sec:bayesian}.

The frequentist confidence interval allows us to avoid making any modeling assumptions (beyond the basic Markov Link Method assumption).  However, it does come with some drawbacks.   For example, we here take the view that values inside the confidence interval are somehow `plausible' or `likely.'  Such a statement cannot be philosophically justified without some at least a hint of Bayesian reasoning (cf.\ \cite{morey2016fallacy}).  It is also important to note that, as is typical for confidence intervals, we make no corrections for multiple testing.  We therefore expect that a small proportion of our confidence intervals will not contain the truth.   It is therefore important to consider all of the intervals produced and be wary of `cherry picking' only those intervals that support a particular conclusion.  However, if cherry picking is desired, a false discovery rate correction may be applied (cf.\ \cite{benjamini2005false}).  A related drawback is that the confidence intervals are not always logically consistent with one another.  For example, it can happen that that $C_{x,y_1}=[0,1]$ and $C_{x,y_2}=[.2,1.0]$.  These intervals don't really make sense together, in that if $q(y_2|x)\geq .2$ then certainly $q(y_2|x)\leq .8$.  A possible remedy would be to construct a simultaneous confidence region for all the parameters concerning a fixed value of $x$.  Analyzing the statistical properties of these high-dimensional confidence regions would be nontrivial; we leave it for future research.

@@

@@

\section{Simulations}

@@

\figgy{
\hfill{}\figgygrs{.4}{../images/simulationI}{(a) Estimation error, \\ varying number of samples \\ and number of sampling procedures.}\hfill{}
\hfill{}\figgygrs{.5}{../images/simulationIconf}{(b) Confidence intervals, \\ varying number of samples \\ and number of sampling procedures.}\hfill{}

\hfill{}\figgygrs{.7}{../images/simulationII}{(c) Estimation error, varying number \\ of samples and strength of link dependency.}\hfill{}

\caption{{\bf When can the MLM accurately estimate the link?} \label{fig:simI}  We use simulations to test the MLM.  In subfigure (a) we look at the error of the MLM link estimator in many different situations, varying the numbers of samples, sampling procedures, and the link parameters.  In each trial our `ground truth' link parameters picked uniformly at random from the parameter space.  The MLM estimator accurately recovers the ground truth from simulated data, as long as we have enough samples \emph{and} enough distinct sampling procedures.  In subfigure (b) we consider the MLM's 95\% confidence intervals.  For a selection of trials we choose a random value of of $x,y$ and then compare the true link parameter $q^*(y|x)$ with the MLM confidence interval $C_{x,y}$.  We show the interval (in black) centered around the ground truth (in red).  Since the link parameters lie in $[0,1]$, the interval is always contained within $\pm1$ of the ground truth (this larger region is shown in gray).  The interval mostly contain the truth, and get narrower with more samples as long as there are enough distinct sampling procedures.   In subfigure (c) we focus on the case of exactly $4$ sampling procedures.  In this case it is only possible to get very accurate estimates if the link takes on a particular form.  If the link is indepenent (i.e.\ $X,Y$ are independent) or if the link is invertible and deterministic (e.g.\ $X=Y$) then we can determine what the link is.  However, if the truth is more complex, small numbers of sampling procedures may make it impossible to determine the true link; this issue is discussed in detail in Section \ref{sec:casestudies}.}
}

We used simulations to test the two parts of the MLM: the point estimators $\tilde q(y|x)$ and confidence intervals $C_{x,y}$.  For any fixed value of the ground truth parameters, $(p^*,q^*)$, we would ordinarily hope for two things:
\begin{description}
    \item[Estimator convergence] As we obtain more samples, $\tilde q(y|x)$ should converge to $q^*(y|x)$ for each $x,y$.
    \item[Conservative coverage] $\mathbb{P}(q^*(y|x) \in C_{x,y})\geq95\%$ for each $x,y$.  That is, for each choice of $x,y$ the confidence interval $C_{x,y}$ should contain $q^*(y|x)$ with probability at least 95\%.\footnote{Note that we are not producing a `simultaneous' interval.  In a simultaneous interval, there would be a 95\% chance that $q^*(y|x)\in C_{x,y}$ for \emph{every} $x,y$.  The key difference is the change from `each' (in our intervals) to `every' (in simultaneous intervals).  Thus, it may be that $q^*(y|x)\notin C_{x,y}$ for a few values of $x,y$.  For \emph{most} values of $x,y$ the true parameters should be contained inside of their corresponding confidence intervals.}
\end{description}
Our simulations yielded good news and bad news.  The good news: the MLM generally achieves conservative coverage.  In many cases, the MLM also achieves estimator convergence.  The bad news: in many cases the MLM fails to achieve estimator convergence.  Indeed, we can show that an `identifiability issue' blocks any possibility of success in some cases; we detail this problem in Appendix \ref{sec:casestudies}.  The following two simulations give a good overview of the relevant issues:
